---
title: "Thesis Methods"
author: "Casey Sakamoto"
date: "6/17/2020"
output: pdf_document
---

```{r setup, include=FALSE}
library(lcmm) # to fit the model 
library(tidyverse) # to mess around w data
library(Hmisc)
# load in analysis dataset (made in prelim-thesis-dataexp rmd)
BONUS_lcmm <- read.csv("C:/Users/Casey/Desktop/WorkData/Thesis 2020/BONUS_lcmm.csv")

# NUMERIC ID
BONUS_lcmm = BONUS_lcmm %>% select(-X) %>% mutate(patient_id = as.numeric(str_sub(Patient,start=2)))
```
NOTE: WE NEED TO CONVERT SUBID TO NUMERIC? IDK WHY BUT SPITTING OUT AN ERROR FOR ME WHEN FITTING MODELS

```{r normality assump}
# clr all groups
hist(BONUS_lcmm$Streptococcus);hist(BONUS_lcmm$Veillonella);hist(BONUS_lcmm$Prevotella);hist(BONUS_lcmm$Neisseria); hist(BONUS_lcmm$Other)
# strep a bit right skew
# veil looks ok
# prev looks like a bit bimodal
# Neis by far the worst
# other pretty right skewed

# ilr all groups (4 cat instead of 5)
hist(BONUS_lcmm$V1); hist(BONUS_lcmm$V2); hist(BONUS_lcmm$V3); hist(BONUS_lcmm$V4)
# v1 doesnt look great
# v2 some left skew
# v3 some right skew
# v4 looks fine
```

```{r data manipulation}
# remove one of B503 age 5 meas (no mb data for the lower one) C02613
# remove subjects with below 2/3 visits al# = at least # visits
# at least 2 visits
BONUS_al2 = BONUS_lcmm %>% filter(!(molecular_id == "C02613" & agemos < 4.8)) %>% filter(Num_Samples > 1)
# at least 3 visits
BONUS_al3 = BONUS_lcmm %>% filter(!(molecular_id == "C02613" & agemos < 4.8)) %>% filter(Num_Samples > 2)
```

```{r prelim models}
# purpose is to see what converges and what doesnt
# g# will be the groups used t# will be the lowest visits allowed per subject inclusion
# _lr will be normalizing method

##### at least 2 data points ######
# all groups included clr, age rounded
g5t2_clr = multlcmm(Streptococcus + Veillonella + Prevotella + Neisseria + Other ~ Age,
                    random = ~1,subject = "patient_id", data = BONUS_al2, maxiter = 300)
summary(g5t2_clr) # doesnt converge up to 300 iter (~ 3.7 min)

# all groups included ilr, age rounded
g4t2_ilr = multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", data = BONUS_al2, maxiter = 300)
summary(g4t2_ilr) # converges!

# excluding Other group clr
g4t2_clr = multlcmm(Streptococcus + Veillonella + Prevotella + Neisseria ~ Age,
                    random = ~1,subject = "patient_id", data = BONUS_al2, maxiter = 500)
summary(g4t2_clr) # doesnt converge up to 500 iter (~ 3 min)

# excluding Neisseria group clr
g4t2_clr_2 = multlcmm(Streptococcus + Veillonella + Prevotella + Other ~ Age,
                    random = ~1,subject = "patient_id", data = BONUS_al2, maxiter = 500)
summary(g4t2_clr_2) # doesnt converge up to 500 iter (~ 3 min)

# excluding Other, Neisseria groups clr
g3t2_clr = multlcmm(Streptococcus + Veillonella + Prevotella ~ Age,
                    random = ~1,subject = "patient_id", data = BONUS_al2, maxiter = 1000)
summary(g3t2_clr) # doesnt converge up to 1000 iter (~ 3 min)

# excluding Other, Prevotella groups clr
g3t2_clr_2 = multlcmm(Streptococcus + Veillonella + Neisseria ~ Age,
                    random = ~1,subject = "patient_id", data = BONUS_al2, maxiter = 1000)
summary(g3t2_clr_2) # doesnt converge up to 1000 iter (~ 3 min)

# CLR Strep + Other
g2t2_clr = multlcmm(Streptococcus + Other ~ Age,
                    random = ~1,subject = "patient_id", data = BONUS_al2, maxiter = 1000)
summary(g2t2_clr) # converges!

# CLR Strep + veil
g2t2_clr_2 = multlcmm(Streptococcus + Veillonella ~ Age,
                    random = ~1,subject = "patient_id", data = BONUS_al2, maxiter = 1000)
summary(g2t2_clr_2) # converges! (~ 3 min)

###### At least 3 data points ########
# all groups included clr, age rounded
g5t3_clr = multlcmm(Streptococcus + Veillonella + Prevotella + Neisseria + Other ~ Age,
                    random = ~1,subject = "patient_id", data = BONUS_al3, maxiter = 300)
summary(g5t3_clr) # doesnt converge up to 300 iter (~ 3.7 min)

# all groups included ilr, age rounded
g4t3_ilr = multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", data = BONUS_al3, maxiter = 300)
summary(g4t3_ilr) # converges!

# clr - other
g4t3_clr = multlcmm(Streptococcus + Veillonella + Prevotella + Neisseria ~ Age,
                    random = ~1,subject = "patient_id", data = BONUS_al3, maxiter = 500)
summary(g4t3_clr)

# clr - other, neisseria
g3t3_clr = multlcmm(Streptococcus + Veillonella + Prevotella ~ Age,
                    random = ~1,subject = "patient_id", data = BONUS_al3, maxiter = 500)
summary(g3t3_clr)

# clr - other, neisseria
g2t3_clr = multlcmm(Streptococcus + Veillonella ~ Age,
                    random = ~1,subject = "patient_id", data = BONUS_al3, maxiter = 500)
summary(g2t3_clr)
```
### Failed Conversion Attempts
#### at least 2 measurements
- All 5 groups CLR
- CLR groups - Other
- CLR groups - Neis
- CLR groups - Other & Neis
- CLR groups - Other & Prev

#### at least 3 measurements

Its looking like ILR works right off the bat, but CLR we need to reduce to only a few (2) groups to get convergence

```{R primary ilr models}
#########################################################################################
# fit different amount of latent classes for ILR set at lest 2 obs per subject
# all groups included

# 1 latent class
mod1_ilr = g4t2_ilr

# 2 latent classes
mod2_ilr = multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 2,B = random(mod1_ilr), data = BONUS_al2)
summary(mod2_ilr) # converges!

mod2_ilr_gs = multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 2,B = random(mod1_ilr), data = BONUS_al2)
summary(mod2_ilr) # converges!
# 3 latent classes
mod3_ilr  = multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 3,B = random(mod1_ilr), data = BONUS_al2)
summary(mod3_ilr) # doesnt converge (but only from the second deriv criteria)

# 4 latent classes
mod4_ilr  = multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 4,B = random(mod1_ilr), data = BONUS_al2)
summary(mod4_ilr) # doesnt converge (but only from the second deriv criteria)

# 5 latent classes
mod5_ilr  = multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 5,B = random(mod1_ilr), data = BONUS_al2)
summary(mod5_ilr) # doesnt converge for all 3 criteria

# 6 latent classes
mod6_ilr  = multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 6,B = random(mod1_ilr), data = BONUS_al2)
summary(mod6_ilr) # doesnt converge (but only from the second deriv criteria)

# 7 latent classes
mod7_ilr  = multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 7,B = random(mod1_ilr), data = BONUS_al2)
summary(mod7_ilr) # doesnt converge (for likleihood and 2nd deriv criteria)

# 8 latent classes
mod8_ilr  = multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 8,B = random(mod1_ilr), data = BONUS_al2)
summary(mod8_ilr) # doesnt converge (but only from the second deriv criteria)

# 9 latent classes
mod9_ilr  = multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 9,B = random(mod1_ilr), data = BONUS_al2)
summary(mod9_ilr) # doesnt converge (for likleihood and 2nd deriv criteria)

# 10 latent classes
mod10_ilr  = multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 10,B = random(mod1_ilr), data = BONUS_al2)
summary(mod10_ilr) # doesnt converge (but only from the second deriv criteria)
#########################################################################################

summarytable(mod1_ilr, mod2_ilr, mod3_ilr, mod4_ilr, mod5_ilr, mod6_ilr, mod7_ilr, mod8_ilr, mod9_ilr, mod10_ilr) # at any point we are only getting 1 or 3 groups
postprob(mod2_ilr) # extremely ambiguous for just 2 latent classes
postprob(mod3_ilr) # extremely ambiguous for 3 lc
```

It seems like the gridsearched initial values classify a lot better; will do that from now on
```{r ilr primary models 3 obs linear}
# fit different amount of latent classes for ILR set at lest 2 obs per subject
# all groups included
# 1 latent class
mod1_ilr3 = g4t3_ilr

# 2 latent classes
mod2_ilr3 = multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 2,B = random(mod1_ilr3), data = BONUS_al3)
summary(mod2_ilr3) # converges!

mod2_ilr3_gs  = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 2, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)

# 3 latent classes
# initial values from model 1
mod3_ilr3  = multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 3,B = random(mod1_ilr3), data = BONUS_al3)
summary(mod3_ilr3) # converges!

# initial values gridsearched
mod3_ilr3_gs  = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 3, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
# 4 latent classes
mod4_ilr3_gs  = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 4, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
#summary(mod4_ilr3_gs) # doesnt converge (but only from the second deriv criteria)

# 5 latent classes
mod5_ilr3_gs  = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 5, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
# summary(mod5_ilr3) # doesnt converge for all 3 criteria

# 6 latent classes
mod6_ilr3_gs  = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 6, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
#summary(mod6_ilr) # doesnt converge (but only from the second deriv criteria)

# 7 latent classes
mod7_ilr3_gs  = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 7, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
#summary(mod7_ilr) # doesnt converge (for likleihood and 2nd deriv criteria)

# 8 latent classes
mod8_ilr3_gs = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 8, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
#summary(mod8_ilr) # doesnt converge (but only from the second deriv criteria)

# 9 latent classes
mod9_ilr3_gs  = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 9, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
#summary(mod9_ilr) # doesnt converge (for likleihood and 2nd deriv criteria)

# 10 latent classes
mod10_ilr3_gs  = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,
                    ng = 10, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
#summary(mod10_ilr) # doesnt converge (but only from the second deriv criteria)

summarytable(mod1_ilr3, mod2_ilr3,mod2_ilr3_gs, mod3_ilr3, mod3_ilr3_gs, mod4_ilr3_gs, mod5_ilr3_gs, mod6_ilr3_gs, mod7_ilr3_gs, mod8_ilr3_gs, mod9_ilr3_gs, mod10_ilr3_gs)
postprob(mod4_ilr3_gs)
summary(mod5_ilr3_gs)
```
splines methods dont converge
```{r models splines 3 obs}
mod1_ilr3_s = multlcmm(V1 + V2 + V3 + V4 ~ Age,link = "5-quant-splines",
                    random = ~1,subject = "patient_id", data = BONUS_al3)
summary(mod1_ilr3_s)

# 2 latent classes
mod2_ilr3_s = multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", 
                    mixture = ~1, link = "5-quant-splines", B = random(mod1_ilr3_s),
                    ng = 2, data = BONUS_al3)

# 3 latent classes
# initial values from model 1
# mod3_ilr3 _ = multlcmm(V1 + V2 + V3 + V4 ~ Age,
#                     random = ~1,subject = "patient_id", mixture = ~1,link = "splines",
#                     ng = 3,B = random(mod1_ilr3), data = BONUS_al3)
# summary(mod3_ilr3) # converges!

# initial values gridsearched
mod3_ilr3_gs_s  = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1, link = "5-quant-splines",
                    ng = 3, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
# 4 latent classes
mod4_ilr3_gs_s  = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1, link = "5-quant-splines",
                    ng = 4, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
#summary(mod4_ilr3_gs) # doesnt converge (but only from the second deriv criteria)

# 5 latent classes
mod5_ilr3_gs_s  = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1, link = "5-quant-splines",
                    ng = 5, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
# summary(mod5_ilr3) # doesnt converge for all 3 criteria

# 6 latent classes
mod6_ilr3_gs_s  = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,link = "5-quant-splines",
                    ng = 6, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
#summary(mod6_ilr) # doesnt converge (but only from the second deriv criteria)

# 7 latent classes
mod7_ilr3_gs_s  = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,link = "5-quant-splines",
                    ng = 7, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
#summary(mod7_ilr) # doesnt converge (for likleihood and 2nd deriv criteria)

# 8 latent classes
mod8_ilr3_gs_s = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,link = "5-quant-splines",
                    ng = 8, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
#summary(mod8_ilr) # doesnt converge (but only from the second deriv criteria)

# 9 latent classes
mod9_ilr3_gs_s  = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,link = "5-quant-splines",
                    ng = 9, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
#summary(mod9_ilr) # doesnt converge (for likleihood and 2nd deriv criteria)

# 10 latent classes
mod10_ilr3_gs_s = gridsearch(multlcmm(V1 + V2 + V3 + V4 ~ Age,
                    random = ~1,subject = "patient_id", mixture = ~1,link = "5-quant-splines",
                    ng = 10, data = BONUS_al3), rep = 30, maxiter = 15, minit = mod1_ilr3)
```

```{r classifications}
# using 1- 4 latent classes
mod2_ilr3$pprob %>% filter(class == 2)
mod3_ilr3_gs$pprob
mod4_ilr3_gs$pprob
# look how subs from 2, 3, 4 latent classes are classified
# first 2 col are the classifications
classed_2 = mod2_ilr3$pprob[,1:2] %>% rename(class2 = class)
classed_3 = mod3_ilr3_gs$pprob[,1:2] %>% rename(class3 = class)
classed_4 = mod4_ilr3_gs$pprob[,1:2] %>% rename(class4 = class)

classifications = full_join(classed_2, classed_3, by = "patient_id")
classifications = full_join(classifications, classed_4, by = "patient_id")

classifications %>% group_by(class2,class3, class4) %>% summarise(n())
```

```{r figures}
plot(mod2_ilr3)
plot(mod2_ilr3)
```

